Pool Model Paper

### 2023-12-06 pool_model.tex
Add citainions to existing models

\newpage
%
%
%
%
% ================================================================================================
\section{Introduction}

Comprehensive reviews about competition/cooperation bacterial mechanisms: \cite{ghoul_ecology_2016, stubbendieck_bacterial_2016, hibbing_bacterial_2010, west_social_2007}
%
%
%
%
% ================================================================================================
\section{Motivating Example}

...

\cmtb{Maybe add here motivations about bacteria states: So toy system where we can eliminate states (sum and get F(n1, n2)).
And describe it initially without states/1state. Get the difference.}

\cmtb{Maybe Example from Seminar by Jens? 2 species,  1 resource, 1 waste production.}
%
%
%
%
% ================================================================================================
\section{General Mathemetical Formulation}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, the system can be described with the Ordinary Differential Equations (\acp{ode}) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
   \label{eq:model_ODE}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector descibing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentartion of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.

The advantage of the following formulation is its generality. 
Depending on the knowledge about the system as well as the needed level of accuracy one can define an arbitrary number of states $M$ and control the complexity of the model.
Moreover, opposite to the already known Predictive Microbiology like Baranyi and Roberts one can easily change.

@@

In the following section we demonstate the application of this approach for different toy models describing the processes within and between bacteria communities.

@@

This so-called preparation to the exponential growth includes synthesys of the needed compounds, initiation of the transciption~\cite{rolfe_lag_2012} and poorly studied physiological and regulatory mechanisms~\cite{monod_growth_1949}
After this, the succeeded in adaptation bacteria enter the exponential growth phase, where the process of cell division is directly happening \cmt{Add Ref}.

@@

Motivated by this we divide a bacterial population into three pools (states) (Figure~\ref{fig:SchematicRep}): 

@@

The suggested model can be cast in the following reactions: \cmt{Use notations x ??}

@@

According to the works of Jacques Monod, in the system where all needed nutritients are in excess except for one the total growth is linear to the concentration of this liminting nutritient~\cite{monod_growth_1949}.
The control of the population density with respect to the available nutritions is implemented using cell communication process called quorum sensing. 

@@

For simplicity we rather make a transition to the dimensionless constant $\alpha:=\alpha_0 N_t$ and diminsionless resource pool $R(t) = R_0 / N_t$ ($0 \leqslant R \leqslant 1$).

@@

$N_t$ denotes the maximal size of the total bacterial population due to environmental limitations (\textbf{Figure \ref{fig:3pool_resource_plots}}). 

@@

To model it we propose a kind of visco-elastic Maxwell type of stress-strain relation and write down the ordinary differential equation for $\gamma$ to be:

@@

    \dot{G}_A &=& \alpha_A G_A\left(1 - \frac{G_A}{N_t}\right) - \frac{\alpha_A}{N_t}G_AG_B
    \label{eq:LV_simple1}\\
    \dot{G}_B &=& \alpha_B G_B\left(1-\frac{G_B}{N_t}\right) -\frac{\alpha_B}{N_t}G_AG_B. 
    \label{eq:LV_simple2}

@@

Equations (\ref{eq:LV_simple1}-\ref{eq:LV_simple2}) are basically competitive Lotka-Volterra equations that are widely used in ecological studies and show the principle of competitive excursion.

@@

\cmt{But LV for several bacteria states in literature? Didn't find!}

@@

This model assumes that toxin is only produced when the total bacterial grow and divide, hence, the total bacterial count $N_B$ increases.\\

@@

    \psi&=&\tilde{\alpha}_B/\tilde{\alpha}_A\\
    \psi&=&\frac{K_B\alpha_B}{K_A\alpha_A}

@@

    \dot{G}_A &=&\frac{G_A}{G_B}\left(1 - G_A-G_B\right)
    \label{eq:2sp_1pool_inhibit_unitless1}\\
    \dot{G}_B &=&\psi\frac{G_B}{G_A}\left(1-G_A-G_B\right). 
    \label{eq:2sp_1pool_inhibit_unitless2}

@@

        The two-species system mutually inhibiting each other is defined with equations~(\ref{eq:2sp_1pool_inhibit_unitless1}-\ref{eq:2sp_1pool_inhibit_unitless2}).

@@

Another mechanism is the motility of bacteria, e.g., the flagellates bacteria as \textit{E. coli} or \textit{B. subtilis} can experience the chemostatic motion of driven by oxygen concentrations~\cite{decoene_microscopic_2011}.

@@

    \item \cite{decoene_microscopic_2011}: 2D chemostatic self-propelled bacterial motion of the flagellated bactreia (e.g. E. coli, B. subtilis).
    Chemostaxis here was driven by the oxygen concentrations.

@@

To test the spatial limitations of the described above approach consider the system consisting from two bacterial species thet can be described bt two internal states: lag and growth states. 

>>>>

\section{Introduction}
Motivation: general, flexible approach for bacterial systems that allows easily modify the system depending on processes on different scales.
Allows us to describe the processes within populations using different bacterial states as well as interactions between different populations using ODEs.\\
Existing models in Predictive Microbiology (Primary Models)~\cite{perez-rodriguez_predictive_2012}:
    \item An overview of primary models used in food science can be found in~\cite{van_boekel_kinetic_2008}.
    \item Baranyi-Roberts Model~\cite{baranyi_modeling_1993, baranyi_dynamic_1994}
    \item modified Gompertz
    \item Buchanan et al.: three-phase linear model for lag phase, exponential growth phase, and stationary phase~\cite{buchanan_when_1997}
    \item McKellar two-compartment model with bacteria in growing and non-growing state~\cite{mckellar_heterogeneous_1997}
    \item Inactivation models
Secondary models:
    \item Polynomial models 
    \item Square Root-Type Models, e.g., Ratkowski for temperature~\cite{ratkowsky_relationship_1982}
    \item The Cardinal Parameter Model~\cite{zwietering_decision_1992}
Comprehensive reviews about competition/cooperation bacterial mechanisms: \cite{ghoul_ecology_2016, stubbendieck_bacterial_2016, hibbing_bacterial_2010, west_social_2007}\\
\section{General Mathematical Formulation}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, the system can be described with the Ordinary Differential Equations (\acp{ode}) using a non-linear function $\mbF$:
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
   \label{eq:model_ODE}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector describing bacterial count.
In matrix form it can be written as
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
    \label{eq:model_bact}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentration of the inhibitors or activators:
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
Here function $\mbH$ is in general case also non-linear.\\
The advantage of the following formulation is its generality.
Depending on the knowledge about the system as well as the needed level of accuracy one can define an arbitrary number of states $M$ and control the complexity of the model.
Moreover, opposite to the already known Predictive Microbiology like Baranyi and Roberts one can easily change.\\

@@

In the following section we demonstrate the application of this approach for different toy models describing the processes within and between bacteria communities.

@@

This so-called preparation to the exponential growth includes synthesys of the needed compounds, initiation of the transciption~\cite{rolfe_lag_2012} and poorly studied physiological and regulatory mechanisms~\cite{monod_growth_1949}.
After this, the succeeded in adaptation bacteria enter the exponential growth phase, where the process of cell division is directly happening.

@@

Motivated by this we divide a bacterial population into three pools (states) (~\ref{fig:SchematicRep}):

@@

The suggested model can be cast in the following reactions:

@@

According to the works of Jacques Monod, in the system where all needed nutritients are in excess except for one the total growth is linear to the concentration of this limiting nutritient~\cite{monod_growth_1949}.
The control of the population density with respect to the available nutrition is implemented using cell communication process called quorum sensing.

@@

For simplicity, we rather make a transition to the dimensionless constant $\alpha:=\alpha_0 N_t$ and dimensionless resource pool $R(t) = R_0 / N_t$ ($0 \leqslant R \leqslant 1$).

@@

$N_t$ denotes the maximal size of the total bacterial population due to environmental limitations (Figure \ref{fig:3pool_resource_plots}).

@@

To model it we propose a kind of viscoelastic Maxwell type of stress-strain relation and write down the ordinary differential equation for $\gamma$ to be:

@@

    \dot{G}_A = \alpha_A G_A\left(1 - \frac{G_A}{N_t}\right) - \frac{\alpha_A}{N_t}G_AG_B\\
    \dot{G}_B = \alpha_B G_B\left(1-\frac{G_B}{N_t}\right) -\frac{\alpha_B}{N_t}G_AG_B. 
    \label{eq:LV_simple}

@@

Equations (\ref{eq:LV_simple}) are basically competitive Lotka-Volterra equations that are widely used in ecological studies and show the principle of competitive excursion.

@@

@@

This model assumes that toxin is only produced when the total bacterial grow and divide, hence, the total bacterial count $N_B$ increases.

@@

    \psi = \tilde{\alpha}_B/\tilde{\alpha}_A = \frac{K_B\alpha_B}{K_A\alpha_A}

@@

    \dot{G}_A =\frac{G_A}{G_B}\left(1 - G_A-G_B\right)\\
    \dot{G}_B =\psi\frac{G_B}{G_A}\left(1-G_A-G_B\right). 
    \label{eq:2sp_1pool_inhibit_unitless}

@@

        The two-species system mutually inhibiting each other is defined with equations~(\ref{eq:2sp_1pool_inhibit_unitless}).

@@

Another mechanism is the motility of bacteria, e.g., the flagellates bacteria as \textit{E. coli} or \textit{B. subtilis} can experience the chemotactic motion of driven by oxygen concentrations~\cite{decoene_microscopic_2011}.

@@

    \item \cite{decoene_microscopic_2011}: 2D chemotactic self-propelled bacterial motion of the flagellated bacteria (e.g. E. coli, B. subtilis).
    Chemotaxis here was driven by the oxygen concentrations.

@@

To test the spatial limitations of the described above approach consider the system consisting from two bacterial species that can be described by two internal states: lag and growth states. 
